Combining global tree cover loss data with historical national forest-cover maps to look at six decades of deforestation and forest fragmentation in Madagascar

### 2017-05-12 deforestmap.Rmd
Finishing discussion

We estimated that natural forests covered 8.9 Mha in Madagascar (15% of the national territory of 587041 km<sup>2</sup>) in 2014 and that Madagascar has lost 44% of its natural forest in six decades since 1953 (37% since 1973). There is an age-old debate [@Burns2016] about the original forest cover in Madagascar to know to which extent humans have altered the natural forest landscape since their large-scale settlement around 800 CE [@Cox2012]. French naturalists Henri Humbert [@Humbert1927] and Henri Perrier de la Bâthie [@Perrier1921] stated that the full island was originally covered by forest, leading to the common statement that 90% of the natural forests have disappeared since the arrival of humans on the island [@Kull2000]. More recent studies counter-balanced that point of view saying that grasslands existed in Madagascar long before human arrival and were determined by climate, natural grazing and other natural factors [@Virah-Sawmy2009, @Vorontsova2017]. Other authors have questioned the entire narrative of extensive alteration of the landscape by early human activity which, through legislation, has severe consequences on local people [@Klein2002; @Kull2000]. Whatever the original proportion of natural forests and grasslands in Madagascar, our results demonstrate that human activities since the 5Os have profoundly impacted the forest landscape and that recent conservation and development programs in Madagascar have failed to stop deforestation in the recent years. Associated to deforestation, we showed that residual forest in Madagascar was highly fragmented with 95% of the forest being at a distance <1.2 km of the forest edge in 2014. Both deforestation and forest fragmentation has strong consequences on biodiversity and carbon emissions in Madagascar. Around 90% of Madagascar's endemic species are forest dependent [@Allnutt2008; @Goodman2005] and @Allnutt2008 estimated that deforestation between 1953 and 2000 had led to an extinction of 9% of the species.

Increased wind speed, vapour pressure deficit and disturbance pressure in the vicinity of forest edges increase tree mortality, resulting in persistent carbon emissions long after deforestation occurs. Fire, wind and desiccation can penetrate up to kilometres into the forest15,19,20, altering microclimate within 100-300m of the forest edge13,21,22,23. Field studies have estimated relative carbon losses in forest edges (e) averaging 11% (ref. 24) and up to 36% (ref. 22) and 50% (ref. 25) of aboveground biomass within the 100m edge zone, and modelling studies have shown biomass reductions of up to 70% in small fragments2

Increase in forest fragmentation has a great impact on both carbon emissions [@Putz2014] and biodiversity loss [@Gibson2013, @Murcia1995].
Biodiversity loss, see @Allnutt2008.

More efforts need to be done to save the few remaining natural tropical forests.

@@

In our study, we have shown that the progressive decrease of the deforestation rate on the period 1990-2005 was followed by a continuous increase in the deforestation rate on the period 2005-2014. We showed that deforestation rate has more than doubled on the period 2010-2014 (98651 ha/yr, 1.08%/yr) compared to 2000-2005 (42296 ha.yr, 0.43%/yr). These results are confirmed by previous studies on deforestation at the national scale for Madagascar [@Harper2007; @MEFT2009; @ONE2015]. Despite differences in the methodologies regarding (i) forest definition (associated to independant visual interpretations of observation polygons to train the classifier), (ii) classification algorithms (Maximum Likelihood Classification for @Harper2007 and @MEFT2009 vs. Random Forests for @ONE2015), (iii) deforestation rate computation method, and (iii) correction for the presence of clouds, these studies gave similar forest-cover estimates and show the same deforestation trend over Madagascar. Our deforestation rate estimates from 1990 to 2014 have been computed from cloud-free maps and can be considered more accurate in comparison with estimates from the previous studies, in particular for the deforestation rates expressed in number of hectares per year. Forest-cover and deforestation rate estimates from this study could be used to update the estimates in the upcoming FAO Forest Resources Assessment reports [@Keenan2015]. Current rates of deforestation for Madagascar (98651 ha/yr, 1.08 %/yr for 2010-2014) could also be used to build reference scenarios for deforestation in Madagascar and help implement deforestation mitigation initatives such as the REDD+ [@Olander2008].

The increase of the deforestation rates after 2005 could be attributed to both population growth and political instability in the country. Nearly 90% of Madagascar's population relies on biomass for their daily energy needs [@Minten2013] and the link between population size and deforestation has previously been demonstrated in Madagascar [@Vieilledent2013, @Gorenflo2011]. With a mean demographic growth rate of about 2.8%/yr and a population which has increased from 16 to 24 million people on the period 2000-2015 [@UN2015], the increasing demand in wood-fuel and space for agriculture is likely to explain the increase in deforestation rates. The political crisis of 2009 [@Ploch2012], followed by several years of political instability and weak governance is also very likely to explain the increase in the deforestation rate observed on the period 2005-2014 [@Smith2003].

>>>>

We estimated that natural forests covered 8.9 Mha in Madagascar (15% of the national territory of 587041 km<sup>2</sup>) in 2014 and that Madagascar has lost 44% of its natural forest in six decades since 1953 (37% since 1973). There is an age-old debate [@Burns2016] about the original forest cover in Madagascar to know to which extent humans have altered the natural forest landscape since their large-scale settlement around 800 CE [@Cox2012]. French naturalists Henri Humbert [@Humbert1927] and Henri Perrier de la Bâthie [@Perrier1921] stated that the full island was originally covered by forest, leading to the common statement that 90% of the natural forests have disappeared since the arrival of humans on the island [@Kull2000]. More recent studies counter-balanced that point of view saying that grasslands existed in Madagascar long before human arrival and were determined by climate, natural grazing and other natural factors [@Virah-Sawmy2009, @Vorontsova2017]. Other authors have questioned the entire narrative of extensive alteration of the landscape by early human activity which, through legislation, has severe consequences on local people [@Klein2002; @Kull2000]. Whatever the original proportion of natural forests and grasslands in Madagascar, our results demonstrate that human activities since the 5Os have profoundly impacted the natural tropical forests and that recent conservation and development programs in Madagascar have failed to stop deforestation in the recent years. Deforestation has strong consequences on biodiversity and carbon emissions in Madagascar. Around 90% of Madagascar's species are forest dependent [@Allnutt2008; @Goodman2005] and @Allnutt2008 estimated that deforestation between 1953 and 2000 had led to an extinction of 9% of the species. The additional deforestation we observed on the period 2000-2014 (around 1Mha of deforested natural forest) worsen this result. Regarding carbon emissions, using the 2010 aboveground forest carbon map by @Vieilledent2016, we estimated that deforestation on the period 2010-2014 has led to the emissions of 40224 Gg (1 Gg = 10<sup>9</sup> g) of carbon in the atmosphere (10055 Gg/yr) and that the residual aboveground forest carbon stock was of 832822 Gg in 2014. Associated to deforestation, we showed that residual forest in Madagascar was highly fragmented with 95% of the forest being at a distance <1.2 km of the forest edge in 2014. Forest fragmentation creates forest fragments too small to maintain viable populations and increases 'edge effects' at forest/non-forest interfaces having great impact on both carbon emissions [@Brinck2017] and biodiversity loss [@Gibson2013; @Murcia1995].

@@

In our study, we have shown that the progressive decrease of the deforestation rate on the period 1990-2005 was followed by a continuous increase in the deforestation rate on the period 2005-2014. We showed that deforestation rate has more than doubled on the period 2010-2014 (98651 ha/yr, 1.08%/yr) compared to 2000-2005 (42296 ha.yr, 0.43%/yr). These results are confirmed by previous studies on deforestation at the national scale for Madagascar [@Harper2007; @MEFT2009; @ONE2015]. Despite differences in the methodologies regarding (i) forest definition (associated to independant visual interpretations of observation polygons to train the classifier), (ii) classification algorithms (Maximum Likelihood Classification for @Harper2007 and @MEFT2009 vs. Random Forests for @ONE2015), (iii) deforestation rate computation method, and (iii) correction for the presence of clouds, these studies gave similar forest-cover estimates and show the same deforestation trend over Madagascar. Our deforestation rate estimates from 1990 to 2014 have been computed from cloud-free maps and can be considered more accurate in comparison with estimates from these previous studies, in particular for the deforestation rates expressed in number of hectares per year. Forest-cover and deforestation rate estimates from this study could be used to update the estimates in the upcoming FAO Forest Resources Assessment reports [@Keenan2015]. Current rates of deforestation for Madagascar (98651 ha/yr, 1.08 %/yr for 2010-2014) could also be used to build reference scenarios for deforestation in Madagascar and help implement deforestation mitigation initatives such as the REDD+ [@Olander2008]. The increase of the deforestation rates we observed after 2005 could be attributed to both population growth and political instability in the country. Nearly 90% of Madagascar's population relies on biomass for their daily energy needs [@Minten2013] and the link between population size and deforestation has previously been demonstrated in Madagascar [@Vieilledent2013; @Gorenflo2011]. With a mean demographic growth rate of about 2.8%/yr and a population which has increased from 16 to 24 million people on the period 2000-2015 [@UN2015], the increasing demand in wood-fuel and space for agriculture is likely to explain the increase in deforestation rates. The political crisis of 2009 [@Ploch2012], followed by several years of political instability and weak governance could also explain the increase in the deforestation rate observed on the period 2005-2014 [@Smith2003]. These results show that despite the conservation policy in Madagascar [@Freudenberger2010], deforestation has dramatically increased at the national level since 2005. More efforts should be done to identify the real causes and mechanisms explaining this rampant deforestation in order to implement new conservation strategies that will be effective at conserving the few remaining natural tropical forests in Madagascar.
